---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
    toc: true
---

```{r include = FALSE}
library(dplyr)
library(ggplot2)
```

## Loading and preprocessing the data
For this analysis I've loaded the *dplyr* and *ggplot2* libraries. To start, I simply read the activity monitoring data from a csv file, with no further processing:
```{r echo = TRUE}
activity_raw <- tbl_df(read.csv('activity/activity.csv'))
print(activity_raw)
```

## What is mean total number of steps taken per day?
One way to explore the dataset is by considering the steps taken each day:
```{r echo = TRUE}
date_summary <- activity_raw %>% group_by(date) %>% 
    summarise(steps = sum(steps))
print(date_summary)
```
From there, I can plot a histogram and find middle values for steps taken per day:
```{r echo = TRUE}
steps_range <- diff(range(date_summary$steps, na.rm = TRUE))
g <- ggplot(date_summary, aes(steps)) 
g + geom_bar(binwidth = steps_range/20) + theme_bw()
```
```{r echo = TRUE}
as.data.frame(summarise(date_summary
    , mean = mean(steps, na.rm = TRUE)
    , median = median(steps, na.rm = TRUE) 
))
```

## What is the average daily activity pattern?
Another useful view is activity during different time intervals throughout the day:
```{r echo = TRUE}
interval_summary <- activity_raw %>% group_by(interval) %>% 
    summarise(avg.steps = mean(steps, na.rm = TRUE))

g <- ggplot(interval_summary, aes(interval, avg.steps)) 
g + geom_line() + labs(x = '5 minute interval', y = 'average steps') +
    theme_bw()
```
From here I can see that the maximum average steps occurs at interval 1705:
```{r echo = TRUE}
as.data.frame(interval_summary[max(interval_summary$avg.steps
                                   , na.rm = TRUE), ])
```


## Imputing missing values
The dataset has many missing values, which can introduce bias. One way to compensate for these missing values is by filling in with the average steps for that interval. 
```{r echo = TRUE}
sum(is.na(activity_raw))
```
```{r echo = TRUE}
processed_df <- activity_raw
processed_df$id <- seq(1:nrow(processed_df))
temp <- tbl_df(merge(processed_df[is.na(processed_df), ], interval_summary, 
                     by = 'interval', all.x = TRUE)) %>% arrange(id)
processed_df[is.na(processed_df), 1] <- temp$avg.steps
processed_df
```
Now I can make a histogram and view middle values as before:
```{r echo = TRUE}
date_summary2 <- processed_df %>% group_by(date) %>% 
    summarise(steps = sum(steps))
g <- ggplot(date_summary2, aes(steps))
g + geom_bar(binwidth = steps_range/20) + theme_bw()
as.data.frame(summarise(date_summary2
    , mean = mean(steps, na.rm = TRUE)
    , median = median(steps, na.rm = TRUE) 
))
```
The histogram has skewed slightly right, and the median has increased slightly. 

## Are there differences in activity patterns between weekdays and weekends?
Since people tend to have different behaviors on weekdays and weekends, we can also see how activity differs on weekdays and weekends:
```{r echo = TRUE}
processed_df$date <- as.Date(processed_df$date)
processed_df$day_descrip <- sapply(processed_df$date, function(x) {
    ifelse(weekdays(x) %in% c('Saturday', 'Sunday'), 'weekend', 'weekday')
})
processed_df
```
```{r echo = TRUE}
interval_summary2 <- processed_df %>% group_by(interval, day_descrip) %>% 
    summarise(avg.steps = mean(steps, na.rm = TRUE))
g <- ggplot(interval_summary2, aes(interval, avg.steps)) 
g + geom_line() + labs(x = '5 minute interval', y = 'average steps') +  
    facet_wrap(~day_descrip, ncol = 1) + theme_bw()
```